---
title: "Checking a data-driven eligibility rule before you trust it"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Checking a data-driven eligibility rule before you trust it}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A rule that relaxes trial eligibility criteria from data promises two things:
more eligible patients, and no loss (indeed a gain) in the treatment-effect
estimate. The two promises need different amounts of data, and only one of them
is cheap to keep. This vignette walks through checking both.

## Set up a cohort

```{r}
library(trialplate)
prep <- tp_prepare(colon_data(), colon_criteria, "trt", "time", "status",
                   colon_ps, tau = 1825)
fit  <- tp_fit(prep)
round(data.frame(criterion = fit$names,
                 dHR_pct = 100 * (exp(fit$phi_HR) - 1),
                 dRMST_days = fit$phi_RMST,
                 cost_pct = 100 * (1 - exp(fit$phi_N))), 1)
```

Two benefit estimands are reported side by side. The hazard ratio is
non-collapsible, so restricting a cohort moves it even when every stratum's
treatment effect is identical; the RMST difference does not have that problem.
Where the two disagree about a criterion, the disagreement is real and has to be
resolved on substantive grounds, not by picking whichever is more convenient.

## Before reading any interaction, check whether one could exist

```{r}
L <- tp_leverage(prep)
c(median = median(L[upper.tri(L)]), max = max(L[upper.tri(L)]))
```

Leverage is computed from eligibility rates alone, before any outcome is touched.
Real protocol criteria are permissive, which drives it towards zero: here no pair
exceeds 0.10. A null interaction in such a matrix is structural, not empirical.
Do not read it as evidence that no effect modification exists.

```{r}
tp_implications(prep)$pairs   # NULL: no criterion implies another
```

## Test the rule out of sample

```{r, eval = FALSE}
M <- tp_rule_eval(prep, R = 500, cores = 2)
g <- function(k) M[, grep(k, colnames(M), fixed = TRUE)[1]]
c(more_patients = mean(g("n_hr")  - g("n_full")  > 0),
  lower_hazard  = mean(g("hr_hr") - g("hr_full") < 0))
#> more_patients  lower_hazard
#>         0.982         0.270
```

The rule is fitted on one half and scored on the other, so neither protocol had
a hand in choosing itself. The patient-count promise holds. The hazard-ratio
promise does not — but that is a statement about this cohort's size, not about
the rule. With 309 patients to learn from, the simulated threshold curve puts
the expected value at 0.279; the observed 0.270 is exactly where it should be.

At roughly 3,000 fitting patients the hazard-ratio promise reaches 0.80, and at
5,167 it reaches 0.94. Confounded assignment, even when correctly adjusted for,
pushes those requirements up by about a factor of two.

## What not to conclude from success

If the effect looks large on your own real-world data, that is not reassurance.
In simulation, omitting one confounder from the propensity model made the
procedure appear as reliable as a randomised design while inflating the hazard
ratio gap it was detecting by 32 percent. Compare the gap under richer and
poorer adjustment: it should *shrink* as adjustment improves.
